---
title: "Complexity of infection (COI)"
author: "Paulo Manrique"
output:
  html_document:
    toc: yes
    number_sections: no
    toc_float: yes
    code_folding: show
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: inline
---

In malaria epidemiology, the term polyclonality represents the
co-occurrence of two or more different parasite clones (e.g.,
haplotypes) within an infected individual. Polyclonality may occur due
to the concurrent transmission of multiple parasite haplotypes from the
same mosquito bite (co-infection) or the acquisition of different
haplotypes through independent bites (super-infection). Both processes
are related to the intensity of transmission. In low transmission
settings, there are few infective bites and therefore little chance for
super-infections or co-infections to occur. Most infections are
therefore monoclonal. On the other hand, when conditions favor an
increase in mosquito prevalence and human-mosquito interaction,
super-infections and co-infections become more likely, generating an
increase in the prevalence of polyclonal infections. Polyclonality rate
is therefore generally considered a positive correlate of malaria
transmission intensity (although other features such as case importation
or relapse behavior in *P. vivax* may modify this relationship).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, include = TRUE)
```


## Differentiating monoclonal and polyclonal infections

There are different ways to differentiate between monoclonal and
polyclonal infections and/or define the number of strains present in a
sample. An easy and widely used way to differentiate between monoclonal
and polyclonal infections is based on the number of heterozygous loci
within the sample ($N_{HetLoci}$),. By using a pre-defined threshold
(e.g., $N_{HetLoci} \ge 1$), samples are classified as polyclonal if
they meet or exceed the threshold and classified as monoclonal if not.
That definition is correct when working with a small number of loci, and
*de-novo* mutations and genotyping errors are not frequent for those
loci. When working with whole-genome or amplicon sequencing
(microhaplotype) information, a sample can contain some heterozygous
loci as a product of the diversification of the parasite population
within the infection (*de-novo* mutations that occurs in the human or in
the mosquito) or due to genotyping errors that occurs during
amplification steps (sWGA, PCR1, or PCR2). For this reason, other
metrics have been proposed to infer polyclonality. One simple metric is
the fraction of heterozygous loci per sample ($p_{Het_{(i)}}$), defined
as:

$$\text{Frac_HetLoci }(p_{Het_{(i)}}) = N_{HetLoci_{(i)}} /N_{(i)}$$$N_{HetLoci_{(i)}}$
represents the number of heterozygous sites for sample $i$ and $N_{(i)}$
represents the total number of observed sites for sample $i$.\
\
Another metric used to infer polyclonality is the within-host divergence
($F_{WS_{(i)}}$) index, defined as follows:

$$F_{WS_{(i)}} = \displaystyle{\sum_{j = 1}^{n}{(1 - \frac{H_{W_j}}{H_{exp_{(j)}}})}}$$

$H_{W_j}$ represents the within-host heterozygosity:

$$H_{W_j} = 1 - \displaystyle{\sum_{a = 1}^{b_j}{(\frac{({rd}_{a}^{j})}{RD_j})^2}}$$

Here, for each sample ($i$), $rd_a^j$ represents the allele $a$ observed
at locus $j$, $b_j$ represents the total number of alleles observed at
locus $j$, and $RD_j$ represents the total read-depth observed at locus
$j$.

$H_{exp_{(j)}}$ is the within-population heterozygosity for locus $j$:

$$H_{exp_{(j)}} = 1 - \displaystyle{\sum{p_j^2}}$$

Here, $p_j$ represents the frequency of each of the alleles observed at
locus $j$ at the population level.

Another way to distinguish between monoclonal and polyclonal samples is
to directly estimate the number of strains in each sample (i.e., define
the 'complexity of the infection' (COI)). Samples with COI equal to 1
are monoclonal, and samples with greater values of COI are polyclonal.
The easiest way to define the COI is to assume that it is equal to the
maximum number of alleles ($\text{max_nAlleles}$) observed at any of the
loci in the sample. More sophisticated approaches based on Markov Chain
Monte Carlo (MCMC) methods also exist to estimate COI from
[biallelic](https://github.com/EPPIcenter/THEREALMcCOIL) and
[multiallelic](https://github.com/EPPIcenter/moire) loci, but these
methods are not currently incorporated in the workflow.

In the current workflow, the user is able to set their own criteria to
distinguish between monoclonal and polyclonal samples based on the
combination of the metrics mentioned above. These criteria are set via
the parameter `poly_formula` (default
`"poly_formula": "NHetLoci>1"`).

1. First load libraries and functions in R:

```{r 'Loading required packages and functions'}
source('~/Documents/Github/MHap-Analysis/docs/functions_and_libraries/amplseq_required_libraries.R')
source('~/Documents/Github/MHap-Analysis/docs/functions_and_libraries/amplseq_functions.R')
```

2. Read the ampseq_object in csv format:

```{r}
ampseq_object = read_ampseq(file = '~/Documents/Github/MHap-Analysis/docs/data/Pviv_example/Pviv_ampseq_filtered2.xlsx', 
                   format = 'excel')
```

3. Identify polyclonal infections by Country

```{r}
poly_by_Country = get_polygenomic(ampseq_object = ampseq_object, 
                                   strata = 'Country',
                                   update_popsummary = FALSE,
                                   na.rm = T,
                                   filters = NULL,
                                   poly_quantile = 0.75,
                                   poly_formula = "NHetLoci > 1")

```
In the following plot you can inspect the distribution of the fraction
of heterozygous loci by sample (Frac_HetLoci), within-host divergence
(Fws), and the maximum number of alleles found in a sample at any locus.

```{r, fig.width=6, fig.height=6, fig.cap = "**Figure 1:** Distribution of the fraction of heterozygous loci by sample (fracHetLoci), within-host divergence (Fws), and the maximum number of alleles found in a sample at any locus. Dotted lines represent the quantiles defined by the poly_quantile argument in Terra."}

poly_by_Country$plot_fracHet_vs_Fws


```

All metrics are shown at the sample level in the following table.
Information specifying which loci show heterozygosity in each sample is
also presented.

```{r}

poly_by_Country$coi_bySample %>% DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))

```

The proportion of samples which show heterozygosity at each locus is
specified below:

```{r}

poly_by_Country$coi_byLoci %>% DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))

```

### Proportion of polyclonal infections by Country

The proportion of polyclonal infections by each category of the
Variable1 is presented in the following table and figure.

```{r}

poly_by_Country$pop_summary %>% DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))

```

Plot the fraction of polyclonal samples per Country

```{r}
plot_poly_by_pop = poly_by_Country$pop_summary %>% 
      ggplot(aes(x = factor(pop, 
                            levels = c(unique(poly_by_Country$pop_summary$pop)[unique(poly_by_Country$pop_summary$pop) != 'Total'], "Total")),
                 y = prop_poly,
                 fill = factor(pop, 
                               levels = c(unique(poly_by_Country$pop_summary$pop)[unique(poly_by_Country$pop_summary$pop) != 'Total'], "Total"))))+
      geom_col(alpha = .6) +
      geom_errorbar(aes(ymin = prop_poly_lower, ymax = prop_poly_upper), width = .2)+
      theme_bw() +
      labs(title = "Frequency of polyclonal infections",
           y = "Frecquency") +
      scale_fill_manual(values = c('firebrick', 'dodgerblue', 'green4', 'gold3', "gray30"))+
      theme(axis.text = element_text(size = 12),
            axis.title = element_blank(),
            legend.position = "none",
            axis.text.x = element_text(angle = 90, vjust = 0.5))
```

```{r}
fig2.height = 2.5 + max(round(nchar(as.character(unique(plot_poly_by_pop$data$pop)))/10, 0))
fig2.width = 1*length(unique(plot_poly_by_pop$data$pop))
```

```{r, fig.width=fig2.width, fig.height=fig2.height, fig.cap="**Figure 2:** Proportions of polyclonal infections by Variable1. 95% confidence intervals are computed using a Fisher exact test."}
plot_poly_by_pop

```


### Add information to the ampseq object and save



```{r}
ampseq_object@metadata = left_join(ampseq_object@metadata,
                                   poly_by_Country$coi_bySample,
                                   by = 'Sample_id'
                                   )

ampseq_object@loci_performance = left_join(ampseq_object@loci_performance,
                                           poly_by_Country$coi_byLoci,
                                           by = join_by(loci == locus)
                                           )

# As an Excel file

write_ampseq(ampseq_object, format = 'excel', name = '~/Documents/Github/MHap-Analysis/docs/data/Pviv_example/Pviv_ampseq_filtered3.xlsx')

# As csv files

write_ampseq(ampseq_object, format = 'csv', name = '~/Documents/Github/MHap-Analysis/docs/data/Pviv_example/Pviv_ampseq_filtered3')

```


## Polyclonal infections in P. falciparum


Read the ampseq_object in csv format:

```{r}
pfal_ampseq_object = read_ampseq(file = '~/Documents/Github/MHap-Analysis/docs/data/Pfal_example/Pfal_ampseq_filtered', 
                   format = 'csv')
```

Identify polyclonal infections by Subnational_level2

```{r}
poly_by_snl2 = get_polygenomic(ampseq_object = pfal_ampseq_object, 
                                   strata = 'Subnational_level2',
                                   update_popsummary = FALSE,
                                   na.rm = T,
                                   filters = NULL,
                                   poly_quantile = 0.75,
                                   poly_formula = "NHetLoci > 1")

```
In the following plot you can inspect the distribution of the fraction
of heterozygous loci by sample (Frac_HetLoci), within-host divergence
(Fws), and the maximum number of alleles found in a sample at any locus.

```{r, fig.width=6, fig.height=6, fig.cap = "**Figure 1:** Distribution of the fraction of heterozygous loci by sample (fracHetLoci), within-host divergence (Fws), and the maximum number of alleles found in a sample at any locus. Dotted lines represent the quantiles defined by the poly_quantile argument in Terra."}

poly_by_snl2$plot_fracHet_vs_Fws


```

All metrics are shown at the sample level in the following table.
Information specifying which loci show heterozygosity in each sample is
also presented.

```{r}

poly_by_snl2$coi_bySample %>% DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))

```

The proportion of samples which show heterozygosity at each locus is
specified below:

```{r}

poly_by_snl2$coi_byLoci %>% DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))

```

### Proportion of polyclonal infections by Subnational_level2 

The proportion of polyclonal infections by each category of the
Variable1 is presented in the following table and figure.

```{r}

poly_by_snl2$pop_summary %>% DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))

```

Plot the fraction of polyclonal samples per Country

```{r}
plot_poly_by_snl2 = poly_by_snl2$pop_summary %>% 
      ggplot(aes(x = factor(pop, 
                            levels = c(unique(poly_by_snl2$pop_summary$pop)[unique(poly_by_snl2$pop_summary$pop) != 'Total'], "Total")),
                 y = prop_poly,
                 fill = factor(pop, 
                               levels = c(unique(poly_by_snl2$pop_summary$pop)[unique(poly_by_snl2$pop_summary$pop) != 'Total'], "Total"))))+
      geom_col(alpha = .6) +
      geom_errorbar(aes(ymin = prop_poly_lower, ymax = prop_poly_upper), width = .2)+
      theme_bw() +
      labs(title = "Frequency of polyclonal infections",
           y = "Frecquency") +
      scale_fill_manual(values = c('firebrick', 'dodgerblue', 'green4', 'gold3', 'orange', "gray30"))+
      theme(axis.text = element_text(size = 12),
            axis.title = element_blank(),
            legend.position = "none",
            axis.text.x = element_text(angle = 90, vjust = 0.5))
```

```{r}
fig2.height = 2.5 + max(round(nchar(as.character(unique(plot_poly_by_snl2$data$pop)))/10, 0))
fig2.width = 1*length(unique(plot_poly_by_snl2$data$pop))
```

```{r, fig.width=fig2.width, fig.height=fig2.height, fig.cap="**Figure 2:** Proportions of polyclonal infections by Variable1. 95% confidence intervals are computed using a Fisher exact test."}
plot_poly_by_snl2

```

### Proportion of polyclonal infections by Subnational_level2 and Quarter_of_Collection

The proportion of polyclonal infections by each category of Variable1
and Variable2 is presented in the following table and figure.

```{r}

pfal_ampseq_object@metadata[['snl2_qc']] = 
  paste(pfal_ampseq_object@metadata[['Subnational_level2']], 
        pfal_ampseq_object@metadata[['Quarter_of_Collection']], sep = '::')
    
poly_by_snl2_qc = get_polygenomic(ampseq_object = pfal_ampseq_object,
                                  strata = "snl2_qc",
                                  update_popsummary = F,
                                  na.rm = TRUE,
                                  filters = NULL,
                                  poly_quantile = .75, 
                                  poly_formula = "NHetLoci > 1"
                                  )    

```

```{r}
poly_by_snl2_qc$pop_summary %>% DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))
```


```{r}

plot_poly_by_snl2_qc = poly_by_snl2_qc$pop_summary %>%
      filter(pop != 'Total')%>%
      mutate(
        Subnational_level2 = stringr::str_split(pop, '::', simplify = TRUE)[,1],
        Quarter_of_Collection = stringr::str_split(pop, '::', simplify = TRUE)[,2],
        prop_poly_lower = case_when(
          prop_poly == 0 ~ 0,
          prop_poly != 0 ~ prop_poly_lower),
        prop_poly_upper = case_when(
          prop_poly == 0 ~ 0,
          prop_poly != 0 ~ prop_poly_upper)
      )%>%
      ggplot(aes(x = Quarter_of_Collection,
                 y = prop_poly,
                 ymin = prop_poly_lower,
                 ymax = prop_poly_upper,
                 fill = Subnational_level2))+
      geom_col()+
      geom_errorbar(width = .2)+
      facet_wrap(~Subnational_level2, ncol = 3)+
      theme_bw()+
      scale_fill_manual(values = c('firebrick', 'dodgerblue', 'green4', 'gold3', 'orange'))+
      labs(title = 'Temporal change of the proportion of polyclonal infections',
           y = "Polyclonal infections",
           x = "Quarter of Collection")+
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
            legend.position =  "none")

```

```{r, include=FALSE}
fig3.height = 3*ceiling(length(unique(plot_poly_by_snl2_qc$data$Subnational_level2))/3)
fig3.width = ceiling(3 *
  0.5*length(unique(plot_poly_by_snl2_qc$data$Quarter_of_Collection)))
```

```{r, fig.width=fig3.width, fig.height=fig3.height, fig.cap="**Figure 3:** Proportions of polyclonal infections by Variable1. 95% confidence intervals are computed using a Fisher exact test."}

plot_poly_by_snl2_qc

```
