---
title: "Genetic Relatedness (IBD) and Connectivity"
author: "Neafsey Lab"
output:
  html_document:
    toc: yes
    number_sections: no
    toc_float: yes
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: inline
---

The genetic relatedness between individual parasite haplotypes and among
parasite populations has several practical uses in the study of malaria.
For example, relatedness information can help determine the geographic
origin of imported infections, define the extent to which parasites are
dispersing or are contained within landscapes, and identify whether
specific strains are being selected for over time. Relatedness
information is also very helpful in understanding longitudinal
(within-individual) infection dynamics. In the case of *P. vivax*, for
example, it can distinguish whether infection represents newly acquired
parasites, recrudescence after treatment, or relapse from longer-lasting
hypnozoite reservoirs. Relatedness information can also help resolve
polyclonality signals, i.e., clarify the number of different haplotypes
co-infecting individual patients.

Relatedness is defined as the probability that, at any locus in the
genome, the alleles sampled from two different individuals are identical
by descent ($IBD$). Genetic markers used for this purpose include SNPs,
microsatellites, and (increasingly) amplicon micro-haplotypes (MHAP).
Relatedness can be estimated using a Hidden Markov Model approach
implemented in the [R package
paneljudge](https://github.com/aimeertaylor/paneljudge/tree/master) (see
mathematical framework in [AR Taylor et al.
2019](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6707449/)). In this
package, relatedness ($r$) is estimated as a function of the haplotype
of the two sampled parasites ($Y^{(i)}$ and $Y^{(j)}$, where $i$ and $j$
denote two different sampled genotypes from the population), the
frequency of the alleles in the population ($f_t(g)$, where $t$ denotes
locus), the physical distance ($d_t$, in base-pairs) between
successively analyzed loci ($t-1$ and $t$), the recombination rate
($\rho$), a switching rate of the Markov chain ($k$), and a constant
genotyping error rate ($\varepsilon$).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, include = TRUE)
```

### Pairwise relatedness comparisons between categories

For this report all possible pairwise IBD comparisons between samples
from different categories of Variable1 and Variable2 are computed, and
the results are shown in the following table:

```{r 'Loading required packages and functions'}
source('~/Documents/Github/intro_to_genomic_surveillance/docs/functions_and_libraries/amplseq_required_libraries.R')
source('~/Documents/Github/intro_to_genomic_surveillance/docs/functions_and_libraries/amplseq_functions.R')
#sourceCpp('~/Documents/Github/intro_to_genomic_surveillance/docs/functions_and_libraries/hmmloglikelihood.cpp')
```

Read the ampseq_object in csv format:

```{r}
ampseq_object = read_ampseq(file = '~/Documents/Github/intro_to_genomic_surveillance/docs/data/Pfal_example/Pfal_ampseq_filtered', 
                   format = 'csv')
```

Run the function `pairwise_hmmIBD`:

```{r}

pairwise_relatedness_table = '~/Documents/Github/intro_to_genomic_surveillance/docs/data/Pfal_example/pairwise_relatedness.csv'

if(!file.exists(pairwise_relatedness_table)){
      pairwise_relatedness = NULL
      
      nChunks = 500
      
      for(w in nChunks){
        start = Sys.time()
        pairwise_relatedness = rbind(pairwise_relatedness,
                                     pairwise_hmmIBD(ampseq_object, parallel = TRUE, w = w, n = nChunks))
        time_diff = Sys.time() - start
        
        print(paste0('step ', w, ' done in ', time_diff, ' secs'))
        
      }
      
      write.csv(pairwise_relatedness,
                '~/Documents/Github/intro_to_genomic_surveillance/docs/data/Pfal_example/pairwise_relatedness.csv',
                quote = FALSE,
                row.names = FALSE)
      
    }else{
      
      pairwise_relatedness = read.csv(pairwise_relatedness_table)
      
    }
```

Plot the distribution of relatedness between sites using the function `plot_relatedness_distribution`

```{r}
plot_relatedness_distribution_between = plot_relatedness_distribution(
      pairwise_relatedness = pairwise_relatedness,
      metadata = ampseq_object@metadata,
      Population = 'Subnational_level2',
      fill_color = rep('gray50', length(unique(ampseq_object@metadata[['Subnational_level2']]))*(length(unique(ampseq_object@metadata[['Subnational_level2']]))-1)/2),
      type_pop_comparison = 'between',
      ncol = 3,
      pop_levels = NULL
    )
```

```{r, eval = FALSE}

View(plot_relatedness_distribution_between$relatedness)

```

```{r, include = FALSE}
fig1.height = 2 * ceiling(length(unique(plot_relatedness_distribution_between$plot$data$Pop_comparison))/3)
```

The distribution of pairwise genetic relatedness values is presented
using histograms as follows:

```{r, fig.width=10, fig.height=fig1.height, fig.cap = "**Figure 1:** Pairwise IBD distribution between categories of Variable1 (panels). The x-axis shows genetic relatedness values, ranging from 0 (unrelated) to 1 (clonal). The y-axis shows the number of pairwise comparisons corresponding to each of these relatedness values. The dotted vertical line represents the median genetic relatedness in the total dataset (including both within and between-population comparisons)."}
plot_relatedness_distribution_between$plot
```

### Fraction of highly related comparisons between categories of Variable1

Based on the defined threshold (0.99),
highly related pairwise comparisons between categories in Subnational_level2 are
counted and proportions with their 95% CI (Fisher exact test) are
calculated. 

```{r}
plot_frac_highly_related_between = plot_frac_highly_related(
      pairwise_relatedness = pairwise_relatedness,
      metadata = ampseq_object@metadata,
      Population = 'Subnational_level2',
      fill_color = rep('gray50', length(unique(ampseq_object@metadata[['Subnational_level2']]))*(length(unique(ampseq_object@metadata[['Subnational_level2']]))-1)/2),
      threshold = 0.99,
      type_pop_comparison = 'between',
      pop_levels = NULL)
```

These values are presented in the following table and in
Figure 2:

```{r, eval = FALSE}
View(plot_frac_highly_related_between$highly_related_table)
```

```{r, include = FALSE}
fig2.height = 2.5 + max(round(nchar(as.character(unique(plot_frac_highly_related_between$plot$data$Pop_comparison)))/10, 0))
fig2.width = 1*length(unique(plot_frac_highly_related_between$plot$data$Pop_comparison))

fig2.width = ifelse(fig2.width > 20, 20, fig2.width)
```

```{r, fig.width=fig2.width, fig.height=fig2.height, fig.cap = "**Figure 2:** Fraction of highly related pairwise comparisons (IBD >= ibd_thres) between categories of Variable1 (x-axis). 95% confidence intervals are computed using a Fisher exact test."}

plot_frac_highly_related_between$plot
```

### Fraction of highly related comparisons between categories of Variable1 over Variable2

Highly related pairwise comparisons between categories in Subnational_level2 and
Quarter_of_Collection are also counted and proportions with their 95% CI (Fisher
exact test) are calculated.

```{r}
plot_frac_highly_related_overtime_between = plot_frac_highly_related_over_time(
        pairwise_relatedness = pairwise_relatedness,
        metadata = ampseq_object@metadata,
        Population = c('Subnational_level2', 'Quarter_of_Collection'),
        fill_color = rep('gray50', length(unique(ampseq_object@metadata[['Subnational_level2']]))*(length(unique(ampseq_object@metadata[['Subnational_level2']]))-1)/2),
        threshold = 0.99,
        type_pop_comparison = 'between',
        ncol = 3,
        pop_levels = NULL)
```

These values are presented in the following
table and in Figures 3-4:

```{r, eval = FALSE}
View(plot_frac_highly_related_overtime_between$frac_highly_related)
```

```{r, include = FALSE}
fig3.height = 0.5*length(unique(plot_frac_highly_related_overtime_between$plot_IBD_correlation_matrix$data$Pop_Date_Yi))

fig3.height = ifelse(fig3.height > 20, 20, fig3.height)

fig3.width = 0.5*length(unique(plot_frac_highly_related_overtime_between$plot_IBD_correlation_matrix$data$Pop_Date_Yj))

fig3.width = ifelse(fig3.width > 20, 20, fig3.width)
```

```{r, fig.width=fig3.width, fig.height=fig3.height, fig.cap = "**Figure 3:** Heatmap matrix of the fraction of highly related pairwise comparisons (IBD >= ibd_thres) between categories of Variable1 and Variable2. The color scale indicates the fraction of highly related samples (increasing with higher intensities of red). The diagonal in the matrix represents pairwise comparisons within categories."}
plot_frac_highly_related_overtime_between$plot_IBD_correlation_matrix
```

```{r, include = FALSE}

fig4.height = 3*ceiling(length(unique(plot_frac_highly_related_overtime_between$plot_frac_highly_related$data$Pop_comparison))/3)

fig4.width = ceiling(3 *
  0.5*length(unique(plot_frac_highly_related_overtime_between$plot_frac_highly_related$data$Date_Yi)))

fig4.width = ifelse(fig4.width > 20, 20, fig4.width)
```

```{r, fig.width=fig4.width, fig.height=fig4.height, fig.cap = "**Figure 4:** Bar plot of the fraction of highly related pairwise comparisons (IBD >= ibd_thres) between categories of Variable1 (panels) and Variable2 (x-axis). 95% confidence intervals are computed using a Fisher exact test."}

plot_frac_highly_related_overtime_between$plot_frac_highly_related
```

Finally, let's create PCoA and network plots summarizing the genetic relatedness
between samples.

```{r}
evectors_IBD = IBD_evectors(ampseq_object = ampseq_object,
                                relatedness_table = pairwise_relatedness,
                                k = length(unique(ampseq_object@metadata$Sample_id)),
                                Pop = 'Subnational_level2', q = 2)

col_vector = brewer.pal(5, 'Accent')

IBD_PCA = evectors_IBD$eigenvector %>% ggplot(aes(x = PC1, y = PC2, color = Subnational_level2))+
      geom_point(alpha = .7, size = 2) +
      stat_ellipse(level = .6)+
      scale_color_manual(values = col_vector)+
      theme_bw()+
      labs(x = paste0('1st PCo (', round(evectors_IBD$contrib[1],1), '%)'),
           y = paste0('2nd PCo (', round(evectors_IBD$contrib[2],1), '%)'),
           color = 'Countries')
```

```{r, fig.width=6, fig.height=4.5, fig.cap="**Figure 5:** Principal coordinate analysis (PCoA) based on the inverse of genetic relatedness (1 - IBD). Colors are randomly assigned based on the categories of Variable1. Each dot represents a sample."}
IBD_PCA
```

```{r, fig.width=6, fig.height=6, fig.cap="**Figure 6:** Network representation of pairwise genetic relatedness (IBD), Colors are randomly assigned based on the categories of Variable1. Each dot represents a sample and the lines connect samples with IBD >= ibd_thres."}


plot_network(pairwise_relatedness,
             threshold = 0.99,
             metadata = ampseq_object@metadata,
             sample_id = 'Sample_id',
             group_by = 'Subnational_level2',
             levels = levels(as.factor(
               ampseq_object@metadata[['Subnational_level2']])),
             colors = col_vector
             )
```
