---
title: "*MHap Analysis Report*: Genetic Relatedness (IBD) and Tranmission"
author: "Neafsey Lab"
output:
  html_document:
    toc: yes
    number_sections: no
    toc_float: yes
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: inline
params:
  RData_image: ""
---

The previous section focused primarily on using relatedness information
for malaria parasite source-tracking and connectivity analysis. An
additional important use case for relatedness information involves the
inference of malaria transmission intensity over space and time. In low
transmission settings, human infections are most often monoclonal (i.e.,
involve only a single parasite strain). Therefore, when parasites
transmit to the mosquito vector and enter the obligate sexual stage of
their life cycle, it is common for self-fertilization to occur between
identical parasite gametes. Parasite propagation thus becomes 'clonal',
i.e., alleles remain fully linked over multiple generations. Low
transmission also makes it more likely for parasite diversity to be lost
via drift, exacerbating the extent of clonality and inbreeding in the
population. In higher transmission settings, on the other hand, higher
parasite prevalence and greater standing parasite genetic diversity make
it more likely for polyclonal infections to occur. The sexual parasite
stage is therefore more likely to involve the co-occurrence of distinct
parasite gametes. This enables effective recombination between
haplotypes and the erosion of highly related parasite relationships.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, include = TRUE)
```

```{r}
library(dplyr)
library(magrittr)
load(params$RData_image)

```

### Pairwise relatedness comparisons within categories of Variable1

For this report all possible pairwise IBD comparisons within samples
from different categories of Variable1 and Variable2 are computed, and
the results are shown in the following table:

```{r}

plot_relatedness_distribution_within$relatedness %>%
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))

```

The distribution of the pairwise genetic relatedness is presented using
histograms as follows:

```{r, include = FALSE}
fig1.height = 2 * ceiling(length(unique(plot_relatedness_distribution_within$plot$data$Pop_comparison))/3)
```

```{r, fig.width=10, fig.height=fig1.height, fig.cap = "**Figure 1:**  Pairwise IBD distribution within categories of Variable1 (panels). The x-axis shows genetic relatedness values, ranging from 0 (unrelated) to 1 (clonal). The y-axis shows the number of pairwise comparisons corresponding to each of these relatedness values. The dotted vertical line represents the median genetic relatedness in the total dataset (including both within and between-population comparisons)."}
plot_relatedness_distribution_within$plot
```

### Fraction of highly related comparisons within categories of Variable1

Based on the threshold defined in the workflow in Terra (`ibd_thres`),
highly related pairwise comparisons within categories in Variable1 are
counted and proportions with their 95% CI (Fisher exact test) are
calculated. These values are presented in the following table and in
Figure 2:

```{r}
plot_frac_highly_related_within$highly_related_table %>%
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))
```

```{r, include = FALSE}
fig2.height = 2.5 + max(round(nchar(as.character(unique(plot_frac_highly_related_within$plot$data$Pop_comparison)))/10, 0))
fig2.width = 1*length(unique(plot_frac_highly_related_within$plot$data$Pop_comparison))
```

```{r, fig.width=fig2.width, fig.height=fig2.height, fig.cap = "**Figure 2:** Fraction of highly related pairwise comparisons (IBD >= ibd_thres) within categories of Variable1 (x-axis). 95% confidence intervals are computed using a Fisher exact test."}
plot_frac_highly_related_within$plot
```

### Fraction of highly related comparisons between categories of Variable1 over Variable2

Highly related pairwise comparisons within categories in Variable1 and
Variable2 are also counted and proportions with their 95% CI (Fisher
exact test) are calculated. These values are presented in the following
table and in Figure 3:

```{r}
if(!is.na(Variable2)){plot_frac_highly_related_overtime_within$frac_highly_related %>%
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))}
```

```{r, include = FALSE}
fig4.height = 3*ceiling(length(unique(plot_frac_highly_related_overtime_within$plot_frac_highly_related$data$Pop_comparison))/3)
fig4.width = ceiling(length(unique(plot_frac_highly_related_overtime_within$plot_frac_highly_related$data$Pop_comparison)) *
  0.5*length(unique(plot_frac_highly_related_overtime_within$plot_frac_highly_related$data$Date_Yi)))

fig4.width = ifelse(fig4.width > 20, 20, fig4.width)
```

```{r, fig.width=fig4.width, fig.height=fig4.height, fig.cap = "**Figure 4:** Fraction of highly related pairwise comparisons (IBD >= ibd_thres) within categories of Variable1 (panles) and Variable2 (x-axis). 95% confidence intervals are computed using a Fisher exact test."}

if(!is.na(Variable2)){plot_frac_highly_related_overtime_within$plot_frac_highly_related}
```
