---
title: 'Performance Report'
author: "Neafsey Lab"
output:
  html_document:
    toc: yes
    number_sections: no
    toc_float: yes
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: inline
params:
  RData_image: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, include = TRUE)
```

```{r 'Molecular surveillance of drug resistance'}
library(dplyr)
library(magrittr)
load(params$RData_image)

```

# Content

1.  Preliminary sequencing yield by Run and Variable1.
2.  Identification of potential genotyping errors.
3.  Read depth yield per sample and locus.
4.  Retention of loci and samples for subsequent analysis.

## Preliminary sequencing yield by Run and Variable1

Sample success was defined as the percentage of samples which amplify a
specific percentage of loci given the allele detection coverage
threshold or number of paired reads that support the allele or ASV in a
sample. By default 6 different coverage threshold have been defined: 1,
5, 10, 20, 50, and 100 paired reads. If user have apply a custom threshold in the
inputs in Terra, it will also appear in each of the following plots.

### Overall sequencing yield

```{r, fig.width=7, fig.height=5, fig.cap = "**Figure 1:** Sample success rate based on different thresholds of minimum read depth coverage per allele (ASV or amplicon sequence variant observed in the sample). x-axis represents the percentage of loci (amplicon or marker) which contains alleles that pass the threshold. y-axis represents the percentage of samples which amplify a specific percentage of loci given the allele detection threshold mentioned above. The plot represents all analyzed samples (no subsetting based on run and additional variables)."}
plot_precentage_of_samples_over_min_abd
```

```{r, include=FALSE}
fig2.height = 3*ceiling(length(unique(plot_precentage_of_samples_over_min_abd_byRun$data$Run))/3)

```

```{r, fig.width=10, fig.height=fig2.height, fig.cap = "**Figure 2:** Sample success rate by Sequencing Run based on different thresholds of minimum read depth coverage per allele (ASV or amplicon sequence variant observed in the sample). x-axis represents the percentage of loci (amplicon or marker) which contains alleles that pass the threshold. y-axis represents the percentage of samples which amplify a specific percentage of loci given the allele detection threshold mentioned above."}
plot_precentage_of_samples_over_min_abd_byRun
```

```{r, include = FALSE}
fig3.height = 3*ceiling(length(unique(plot_precentage_of_samples_over_min_abd_byVariable1$data$Variable1))/3)
```

```{r, fig.width=10, fig.height=fig3.height, fig.cap="**Figure 3:** Sample success rate by Sequencing Run based on different thresholds of minimum read depth coverage per allele (ASV or amplicon sequence variant observed in the sample). x-axis represents the percentage of loci (amplicon or marker) which contains alleles that pass the threshold. y-axis represents the percentage of samples which amplify a specific percentage of loci given the allele detection threshold mentioned above."}
plot_precentage_of_samples_over_min_abd_byVariable1
```

## Identifying likely genotyping errors

After the denoising process of the fastq files, the cigar table might 
contains some artefacts that affect downstream analysis.  These artifacts 
include systematic PCR errors (off-target products and stutters in 
INDELs) as well as stochastic PCR errors. All these artifacts 
coexist with the true allele within the sample, causing the observation of 
two or more alleles (the true allele plus the artifacts) in the particular 
loci of the sample where the error has occurred, the increment of the 
heterozygousity of the loci (if the error is systematic), and the increment
of polyclonal infections in the population. 

However, when analyzing a population we do not expect all samples to be
polyclonal, and it should not be possible to find a locus that is
heterozygous for all samples. Moreover, loci with alternative alleles
only present in heterozygous samples should be rare as they are the
result of *"de-novo"* mutations, and the frequency of heterozygous
samples for that loci should be low because samples comes from different
regions and we do not expect a wide geographic distribution of that
genotype (This might not be true if only one population is been
analyzed).

Here the identification and removal of likely genotyping errors is done in
three sequential steps: 1) Removal of off-target products, 2) masking of 
INDELs present in flanking regions of the ASV, 3) and removal of stochastic 
PCR errors (alleles mainly present as minor alleles).

The identification of all these likely genotyping errors is based on the
following parameters:

1.  dVSITES_ij: Density of variant sites (SNPs and INDELs) in the allele (ASV) 
    $j$ of the locus $i$.

2.  P_ij ($P_{i,j}$): The total number of samples that amplified each
    alternative allele $j$ in locus $i$.

3.  H_ij ($H_{i,j}$): Number of heterozygous samples in the locus
    $i$ that carry the alternative allele $j$.

4.  H_ijminor ($H_{i,jminor}$): Number of heterozygous samples in the
    locus $i$ where the alternative allele $j$ is the minor
    Allele (the allele with the lower read counts).

5.  h_ij ($h_{i,j}=H_{i,j}/P_{i,j}$): ratio of heterozygous samples in the 
    locus that carry the alternative allele of interest respect to the
    total number of samples that amplified alternative allele.

6.  h_ijminor ($h_{i,jminor}=H_{i,jminor}/H_{i,j}$): ratio of heterozygous samples
    where the alternative allele is the minor Allele respect to the
    total number of heterozygous samples in the locus that carry
    the alternative allele.

7.  p_ij ($p_{i,j}$): population prevalence of the alternative allele $j$ in
    locus $i$.
    
8.  flanking_INDEL: Boolean in indicating the presence of INDELs in the flanking region of the ASV.

9.  SNV_in_homopolymer: Boolean in indicating the presence of a SNV in an homopolymer region in the ASV.

10. INDL_in_homopolymer: Boolean in indicating the presence of an INDEL in an homopolymer region in the ASV.

### Identification and removal of off-target products

A non-specific product is the result of the amplification of a DNA template 
that shares a certain degree of identity in the region of the primers with 
the desired target. However, the degree of identity of these off-targets with 
respect to the desired target tends to be low, generating a high density of 
polymorphisms in the alignment between the off-target ASV and the reference 
template. For that reason in this step we use the density of of variant sites 
(dVSITES_ij) in order to identify alleles that are potentially off-target 
products. The other parameters explained above can also be included to 
constraint more the definition. The filtering criteria can be specified in 
Terra with the argument "off_target_formula" (by default 
`"off_target_formula": "dVSITES_ij>=0.3"`). 

```{r}
print(paste0(n_off_target_alleles, ' allele(s) match(es) the criteria to define off-target products'))
```

```{r, fig.width=8, fig.height=4, fig.cap="**Figure 4:** Identifcation of off-target products. Histogram represents the distribution of the densitity of variant sites in each allele in each locus (dVSITES_ij). The scatter plot shows the distribution of the prevalence of the alternative allele (p_ij), the frequency of the alternative allele as heterozygous (h_ij), the frequency of the alternative allele as minor allele (h_ijminor), and the density of variant sites (dVSITES_ij) of the alternative allele respect to the reference template."}
plot_off_target_stats
```

```{r}
if(length(n_off_target_alleles) > 0){
  off_target_stats%>%
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))
}
```

### Products with flanking INDELs

Flanking INDELs are defined as INDELs that occurs at the beginning or
the end of the ASV attached to the primer area. Flanking INDELs can lead
to primer miss binding during PCR by creating additional alleles which
differs in the length of the INDEL in the same sample.
The filtering criteria can be specified in Terra with the argument
"flanking_INDEL_formula" (by default 
`"flanking_INDEL_formula": "flanking_INDEL==TRUE&h_ij>=0.66"`). 

Identified Flanking INDELs are masked, which means that the internal 
region of the ASV is kept to define the allele.

```{r}
print(paste0(n_flanking_INDEL_alleles, ' allele(s) match(es) the criteria to identify products with flanking INDELs'))
```

```{r, fig.width=7, fig.height=7, fig.cap="**Figure 5:** Identification of flanking INDELs. The scatter plot shows the distribution of the prevalence of the alternative allele (p_ij), the frequency of the alternative allele as heterozygous (h_ij), and the frequency of the alternative allele as minor allele (h_ijminor). The panels FALSE and TRUE represent alleles that do not contain or contain flanking INDELs."}
plot_flanking_INDEL_stats
```

```{r}
if(n_flanking_INDEL_alleles > 0){
  flanking_INDEL_stats%>%
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))
}
```

### Products with SNVs in homopolymer regions

Homopolymers are regions of low complexity formed by the tandem repetition of the same nucleotide. In these regions, DNA polymerase tends to make stuttering errors, generating INDELs or erroneous incorporation of nucleotides (SNVs). In this section, SNVs and INDELs in homopolymer regions is identified, and their frequency in samples (p_ij), in heterozygous samples (h_ij), and as a minor allele (h_ijminor) is described.

For detecting and masking SNVs in homopolymer regions the filtering criteria can be specified in Terra with the argument
"SNV_in_homopolymer_formula" (by default 
`"SNV_in_homopolymer_formula": "SNV_in_homopolymer==TRUE"`). 

Identified SNVs in homopolymer regions are masked, which means that the other 
regions of the ASV are kept to define the cigar string of the allele.

```{r}
print(paste0(n_SNV_in_homopolymer_alleles, ' allele(s) match(es) the criteria to identify products with SNVs in homopolymers'))
```

```{r, fig.width=7, fig.height=7, fig.cap="**Figure 6:** Identification of SNVs in homopolymers. The scatter plot shows the distribution of the prevalence of the alternative allele (p_ij), the frequency of the alternative allele as heterozygous (h_ij), and the frequency of the alternative allele as minor allele (h_ijminor). The panels FALSE and TRUE represent alleles that do not contain or contain SNVs in homopolymers."}
plot_SNV_in_homopolymer_stats
```

```{r}
if(n_SNV_in_homopolymer_alleles > 0){
  SNV_in_homopolymer_stats%>%
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))
}
```

### Products with INDELs in homopolymer regions

For detecting and masking INDELs in homopolymer regions the filtering criteria can be specified in Terra with the argument
"INDEL_in_homopolymer_formula" (by default 
`"INDEL_in_homopolymer_formula": "INDEL_in_homopolymer==TRUE"`). 

Identified INDELs in homopolymer regions are masked, which means that the other 
regions of the ASV are kept to define the cigar string of the allele.

```{r}
print(paste0(n_INDEL_in_homopolymer_alleles, ' allele(s) match(es) the criteria to identify products with INDEL in homopolymers'))
```

```{r, fig.width=7, fig.height=7, fig.cap="**Figure 7:** Identification of INDELs in homopolymers. The scatter plot shows the distribution of the prevalence of the alternative allele (p_ij), the frequency of the alternative allele as heterozygous (h_ij), and the frequency of the alternative allele as minor allele (h_ijminor). The panels FALSE and TRUE represent alleles that do not contain or contain INDELs in homopolymers."}
plot_INDEL_in_homopolymer_stats
```

```{r}
if(n_INDEL_in_homopolymer_alleles > 0){
  INDEL_in_homopolymer_stats %>%
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))
}
```

### Products detected as potential bimeras

A bimera is a single DNA sequence originating when multiple DNA sequences get joined during PCR amplification. Bimeras are artifacts and should be filtered out from the data during processing to prevent spurious inferences of biological variation.

The filtering criteria can be specified in Terra with the argument
"bimera_formula" (by default 
`"bimera_formula": "bimera==TRUE"`). 

Identified bimeras are removed, which means that the ASV and its cigar string is not considered in further analysis.

```{r}
print(paste0(length(n_flanking_INDEL_alleles), ' allele(s) match(es) the criteria to identify products with flanking INDELs'))
```

```{r, fig.width=7, fig.height=7, fig.cap="**Figure 8:** Identification of bimeras. The scatter plot shows the distribution of the prevalence of the alternative allele (p_ij), the frequency of the alternative allele as heterozygous (h_ij), and the frequency of the alternative allele as minor allele (h_ijminor). The panels FALSE and TRUE represent alleles that are bimeras."}
plot_bimera_stats
```

```{r}
if(n_bimera_alleles > 0){
  bimera_stats%>%
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))
}
```

```{r}
if(n_bimera_alleles > 0){
  bimera_stats%>%
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))
}
```

### Stochastic PCR errors

PCR errors introduce new alleles in both polymorphic and monomorphic
sites. These errors depend on the error rate of the polymerases used in
the sWGA and library generation steps, and they will generate
alternative alleles that are mainly present as the minor allele at a
heterozygous site and their frequency tends to be low in the population.
By analyzing the distribution of population frequency of each
alternative allele p_ij ($p_{i,j}$), the ratio of heterozygous samples in the
locus that carry the alternative allele of interest respect to
the total number of samples that amplified alternative allele h_ij
($h_{i,j}$), and the ratio of heterozygous samples where the alternative
allele is the minor allele respect to the total number of heterozygous
samples h_ijminor ($h_{i,jminor}$), in this step we will remove all alternative
alleles that match our filtering criteria. This filtering criteria can
be specified in Terra with the argument "PCR_errors_formula" (by
default "PCR_errors_formula": `"h_ij>=0.66&h_ijminor>= 0.66"`)

```{r}
print(paste0(n_PCR_errors_alleles, ' allele(s) match(es) the criteria to identify PCR_errors'))
```

```{r, fig.width=7, fig.height=5, fig.cap="**Figure 9:** Identification and removal of stochastic PCR errors. The scatter plot shows the distribution of the prevalence of the alternative allele (p_ij), the frequency of the alternative allele as heterozygous (h_ij), and the frequency of the alternative allele as minor allele (h_ijminor). The horizontal panels are defined based on the threshold stablished for h_ijminor."}
plot_PCR_errors_stats
```

```{r}
if(n_PCR_errors_alleles > 0){
  PCR_errors_stats%>%
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))
}
```

## Read depth yield per sample and locus

```{r, include = FALSE}
Fig5.height = length(unique(ReadDepth_coverage$plot_read_depth_heatmap$data$var))*3
```

```{r, fig.width=12, fig.height=Fig5.height, fig.cap="**Figure 10:** Read Coverage per sample per locus. The heatmaps show the number of read-pairs obtained per sample (rows) at each locus (columns). Separate panels are used for each value of Variable1 (e.g., sample geographic origin). Darker reds indicate higher read-depth. Grey indicates lack of signal for the locus/sample."}
ReadDepth_coverage$plot_read_depth_heatmap
```

```{r, include = FALSE}
Fig6.height = length(unique(ReadDepth_coverage_by_run$plot_read_depth_heatmap$data$var))*3
```

```{r, fig.width=12, fig.height=Fig6.height, fig.cap="**Figure 11:** Read Coverage per sample per locus by run. The heatmaps show the number of read-pairs obtained per sample (rows) at each locus (columns). Separate panels are used for each sequencing run. Darker reds indicate higher read-depth. Grey indicates lack of signal for the locus/sample."}
ReadDepth_coverage_by_run$plot_read_depth_heatmap
```

```{r}
Fig7.height = length(unique(ReadDepth_coverage_by_run_controls$plot_read_depth_heatmap$data$Sample_id))*0.1
```

```{r, fig.width=12, fig.height=10, fig.cap="**Figure 12:** Read Coverage for the controls in each sequencing run. The heatmaps show the number of read-pairs obtained per control (rows) at each locus (columns). Separate panels are used for each sequencing run. Darker reds indicate higher read-depth. Grey indicates lack of signal for the locus/sample. The name of positive controls indicates the strain that has been used and the tag 'pos' (e.g. dd2_pos2), while negative controls has the tag neg with the prefix that indicates the step on which it was incorporated to the plate (EXTR, PCR or sWGA)."}
ReadDepth_coverage_by_run_controls$plot_read_depth_heatmap
```

## Retention of loci and samples for subsequent analysis

Loci and samples with low amplicfication success will be discarded based on 
the locus_ampl_rate and samp_ampl_rate parameter defined in Terra.

### Locus performance across different categories for Variable1

```{r}
fig8.height = 2*length(unique(all_loci_amplification_rate$data$Strata))
```

```{r, fig.width=7, fig.height=fig8.height, fig.cap="**Figure 13:** Locus amplification rate distribution across categories in Variable1. Each panel represents the cateories of Variable1, and the bottom panel represents the total population and the actual number of loci that are retained. Vertical line represents the locus_ampl_rate threshold set by the user in Terra. Only "}

all_loci_amplification_rate

```

### Retention of samples for subsequent analysis

```{r}
fig9.height = 2*length(unique(samples_amplification_rate$data$Strata))
```

```{r, fig.width=7, fig.height=fig9.height, fig.cap="**Figure 14:** Sample amplification rate distribution across categories in Variable1. Each panel represents the categories of Variable1 and the number within each panel represents the number of samples that are retained in each category."}

samples_amplification_rate

```

## Exportable tables

**Table 1: Cigar table without masking and filter**
```{r}
cigar_table_unmasked_unfiltered %>%
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))
```

**Table 2: Masked and filtered cigar table without controls**
```{r}
cigar_table_masked_filtered %>%
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                  buttons = c('csv', 'excel')))
```

**Table 3: Masked and filtered cigar table of controls**
```{r}
cigar_table_controls_masked_filtered %>%
  DT::datatable(extensions = 'Buttons',
                options = list(
                  buttons = c('csv', 'excel')))
```

**Table 4: Metadata of kept samples**
```{r}
metadata_kept_samples %>%
  DT::datatable(extensions = 'Buttons',
                options = list(
                  buttons = c('csv', 'excel')))
```

**Table 5: Metadata of removed samples**
```{r}
metadata_removed_samples %>%
  DT::datatable(extensions = 'Buttons',
                options = list(
                  buttons = c('csv', 'excel')))
```
